# Introduction {#intro}

Welcome to our guide!

From big data to machine learning, as data science becomes an area of ever-increasing importance, it is essential for people to gain an understanding of the foundational theory and applications of data science. Whether it's for a job, freelancing, or personal fun, our aim is to teach you the basics so you can make beautiful data representations like this:

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

For this guide, we'll be using R, and our dataset will be Kaggle's global terrorism database, which can be accessed from here: https://www.kaggle.com/START-UMD/gtd?select=globalterrorismdb_0718dist.csv.

# Getting Started

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
