# Introduction {#intro}

Welcome to our guide!

From big data to machine learning, as data science becomes an area of ever-increasing importance, it is essential for people to gain an understanding of the foundational theory and applications of data science. Whether it's for a job, freelancing, or personal fun, our aim is to teach you the basics so you can make beautiful data representations like this:

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

For this guide, we'll be using R, and our dataset will be Kaggle's global terrorism database, which can be accessed from here: https://www.kaggle.com/START-UMD/gtd?select=globalterrorismdb_0718dist.csv.

# Getting Started

First, put the tidyverse library into your R file.

```{r lib}
library(tidyverse)
```

This library comes with a variety of functions and operators to make data manipulation and tidying easier. For a comprehensive overview, see here: https://medium.com/@brianward1428/introduction-to-tidyverse-7b3dbf2337d5.

